---
title: 'Exploration of media and trust before the campaign'
author: "Wouter van Atteveldt, Nel Ruigrok, Mariken van der Velden"
permalink: reports/prewave_exploration.html
output: 
  md_document:
    preserve_yaml: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo=F, fig.path='', fig.width = 10)
library(printr)
options(digits = 2)
```


This report gives an overview of the political and media landscape at the start of the campaign for the Dutch 2021 Parliamentary Elections. 
This analysis is based on the first wave of our panel survey with in total 2.400 respondents. 
We asked them what their vote intention is for this election, how they voted in the previous election,
how well they trust various media and institutions, and what media they consumed in the past weeks. 

```{r "Read Data"}
library(here)
library(tidyverse)
source(here("src/lib/data.R"))
d_raw <- read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv"))
resp = extract_wide(d_raw) %>%  mutate(age2=case_when(age %in% c("<24" ,"25-34") ~ "<35",
                        age %in% c("55-64", ">64") ~ "55+",
                        T ~ "35-55")) 
long = extract_long(d_raw)
long_vote = resp %>% select(iisID, vote) %>% inner_join(long)
theme_nrc = function(...) ggthemes::theme_hc(...) + theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x=element_blank(), axis.title.y=element_blank())
```

## Media use by Age and Education

```{r}
overall = long %>% filter(variable=="I1") %>% rename(medium=name) %>% inner_join(resp)
news_agg = overall %>% group_by(age2, education, medium) %>% summarize(value=mean(value), n=n())
news_agg %>% filter(education!="Don't know") %>% mutate(medium=fct_reorder(medium, value)) %>% 
  ggplot(aes(x=education, y=age2, label=round(value, 1), fill=value)) + 
  geom_tile()+ geom_text()+
  facet_wrap("medium") + 
  ggtitle("Overall use of different media per age and education group", "(in days per week)")  + 
  xlab("Education group (high=University/HBO, med=HAVO/MBO2+, low=MAVO/MBO1") + ylab("Age group") +
  scale_fill_gradient(low="white", high="#3182bd", guide=F)
```

As we can see, Television is the most widely consumed news channel for all age and education groups,
but with a clear smaller share for respondents under 35, except for those in the lowest education group. 
Unsurprisingly, print newspapers ad opinion magazines are mostly consumed by highly educated older voters,
while younger voters prefer online news channels, especially the highly educated voters younger than 55.
Social media and news apps are seen as a news source especially for young people, regardless of education. 

# Political Shifts

## Vote intention per Age and Education

```{r}
d = resp %>% select(iisID, age2, education, vote) %>% na.omit() %>% 
  group_by(age2, education, vote) %>% summarize(n=n()) %>% group_by(vote) %>% mutate(zetels=n/sum(n)*100) 
ggplot(d, aes(x=education, y=age2, label=round(zetels), fill=zetels)) +
  geom_tile()+ geom_text()+
  facet_wrap("vote") + 
  ggtitle("Vote intention per age and education group", "(in percentage of age/education group)")  + 
  xlab("Education group (high=University/HBO, med=HAVO/MBO2+, low=MAVO/MBO1") + ylab("Age group") +
  scale_fill_gradient(low="white", high="#3182bd", guide=F)
```

The figure above shows how the age and education for the respondents planning to vote for each party.
There are some clear patterns.
On the right, the VVD mostly appeals to young and highly educated voters,
while CDA and PVV are more popular with older and less highly educated voters.
On the left, D66 and GroenLinks appeal to the young and educated class,
while PVdA apd SP appeal more to the older and less educated voter. 

Placing each party by the demographic of its mean voter, we get an alternatieve electoral compass:
(color indicating average position of respondent on left-right scale)

```{r}
library(ggthemes)
resp %>% select(iisID, age, education, vote, rile_self)  %>% 
  filter(!is.na(education), !is.na(vote)) %>% 
  mutate(education=3-as.numeric(education), age=as.numeric(age)) %>% 
  mutate(education = (education - mean(education)) / sd(education), age=(age - mean(age)) / sd(age)) %>% 
  group_by(vote) %>% summarize(n=n(), age=mean(age), education=mean(education), rile=mean(rile_self, na.rm=T)) %>% 
  ggplot(aes(x=education, y=age, label=vote, color=rile)) + geom_label() + theme_void() + theme(axis.title = element_text()) +
  scale_color_gradient2(low="red", mid="black", high="blue", midpoint=5)

```


```{r eval=F}
library(igraph)
resp
ntot = votes %>% filter(A2 != "Weet niet", E1 != "Weet niet") %>% nrow()
gdf = votes %>% filter(A2 != "Weet niet", E1 != "Weet niet") %>% 
  group_by(A2, E1) %>% summarize(n=n()) %>% na.omit() %>% filter(A2 != E1)

gd = full_join(gdf, gdf, by=c("A2"="E1", "E1"="A2")) %>% replace_na(list("n.x"=0, "n.y"=0)) %>% mutate(n=n.x - n.y, zetels=n/ntot * 150) %>% filter(n>0)
g = gd %>% select(E1, A2, n, zetels) %>% filter(n>5) %>% graph_from_data_frame()

data$netwerk_veranderingen = gd

E(g)$width = 1+E(g)$n/10 
E(g)$label = round(E(g)$zetels, 1)
E(g)$arrow.size=E(g)$width/2

plot(g)
```


## Media Use by Vote Intention


```{r aux}
# Overall ordering of parties by institutional trust
parties = long_vote %>% filter(variable=="F2") %>% group_by(vote) %>% 
  summarize(value=mean(value)) %>% arrange(value) %>% pull(vote)
parties = c("Niet", "Anders", setdiff(parties, c("Niet", "Anders")), "Gemiddeld")
n_party = resp %>% group_by(vote) %>% summarize(ntot=n()) %>% na.omit()
ntotal = sum(n_party$ntot)
add_zeroes= function(x) x %>% pivot_wider(values_fill=0) %>% pivot_longer(-vote)

# Helper functions
create_table = function(long_data, compute_perc=FALSE) {
  d = long_data %>% group_by(vote, name) %>% summarize(n=n(), value=mean(value, na.rm=T)) %>%
    inner_join(n_party) %>% mutate(value=if (compute_perc) n/ntot*100 else value)
  total = long_data %>% group_by(name) %>% summarize(n=n(), value=mean(value, na.rm=T)) %>% 
    mutate(value=if (compute_perc) n/ntotal*100 else value) %>% 
    add_column(vote="Gemiddeld") 
  bind_rows(d, total) %>% select(vote, name, value) %>% 
    ungroup() %>% add_zeroes() %>% 
    mutate(
      vote = factor(vote, levels=parties),
      name=fct_reorder(name, value, mean)
    )
}
color_table_plot = function(data, title, subtitle=NULL, digits=0, print=T, export=T) {
  plot = ggplot(data, aes(x=vote, y=name, label=round(value, digits = digits), fill=value)) +
    geom_tile()+ geom_text()+
    ggtitle(title, subtitle)  + 
    scale_fill_gradient(low="white", high="#3182bd", guide=F) +
    theme_nrc()
  if (print) print(plot)
  if (export) data %>% 
    pivot_wider(names_from="vote", names_sort=T) %>% 
    export_data(title) %>% str_c("Download data: ", .) %>% cat()
  if (print) invisible(plot) else plot
}
```

```{r wave0-media-party, results='asis'}
d = long_vote %>% filter(variable=="I1") %>% create_table()
color_table_plot(d, "Wave 0: Vote intention and media use", digits=1, export=F)
```

## Use of particular channels {.tabset}

```{r wave0-media-party-specific, results='asis'}
pct_variables = c(I2="Newspapers", I3="TV", I4="Online", I5="Social", I7="Apps")
for (var in names(pct_variables)) {
  cat(str_c("\n\n### Use of ", pct_variables[[var]], "\n\n"))
  label = paste("Wave 0: Vote intention and news: ", pct_variables[[var]])
  message(paste(var, label))
  d = long_vote %>% filter(variable==var, name != "Other, namely:", value != 0) %>% create_table(compute_perc = T)
  color_table_plot(data=d, title = label, subtitle = "(Percentage of voters for that party; multiple options possible)")
}
```

# Trust

## Trust in Media

```{r wave0-trust-media, results='asis'}
d = long_vote %>% filter(variable=="I8") %>% create_table()
color_table_plot(d, "Wave 0: Trust in Media", subtitle="(Grade from 0=low tot 10=high)", digits=1)
```

## Trust in Institutions

```{r wave0-trust-institution, results='asis'}
d = long_vote %>% filter(variable=="F2") %>% create_table()
color_table_plot(d, "Wave 0: Trust in Instituties", subtitle="(Grade from 0=low tot 10=high)", digits=1)
```

```{r wave0-trust-group, results='asis'}
d = long_vote %>% filter(variable=="F2") %>%
  mutate(name=case_when(
    name == "Science" ~ "Wetenschap",
    name == "The justice system" ~ "Rechtspraak",
    name == "Dutch democracy" ~ "Democratie",
    name == "Journalism" ~ "Journalistiek",
    name %in% c("Banks", "Big corporations") ~ "Bedrijfsleven",
    T ~ "Politiek"
  ))%>% create_table()
color_table_plot(d, "Wave 0: Trust in Institutions (grouped)", 
                    subtitle="(Grade from 0=low tot 10=high)", digits=1)
```

## Trust in democracy

```{r wave0-democracy, results='asis'}
d = long_vote %>% filter(variable=="F2", name == "Dutch democracy") %>% mutate(ok=value>5.5)
x = d %>% group_by(ok, value) %>% summarize(n=n()) %>% ungroup() %>% mutate(perc=n/sum(n)*100)
ggplot(x) + geom_col(aes(x=value, y=perc, fill=ok)) + scale_x_continuous(breaks=0:10) + 
  scale_fill_manual(values=c(scales::muted("red"), scales::muted("green")), guide=F) +
  theme_nrc() + ggtitle("Rapportcijfer voor Democratie")
```

```{r wave0-democracy2, results='asis'}
dt = d %>% group_by(vote, value) %>% summarize(n=n()) %>% mutate(perc=n/sum(n)*100)

total = d %>% group_by(value) %>% summarize(n=n()) %>% mutate(perc=n/sum(n) * 100) %>% 
    add_column(vote="Gemiddeld") 
dt = bind_rows(dt, total) %>% select(vote, name=value, value=perc) %>% 
  ungroup() %>% add_zeroes() %>% 
  mutate(vote = factor(vote, levels=parties),
         name=factor(as.character(name), levels=as.character(0:10)))

color_table_plot(dt, "Wave 0: Trust in democracy", subtitle="(Percentage of voters for a party)")
```
