---
title: 'Data Cleaning: Dutch Parliamentary Elections 2021 Wave 1'
author: "VU Political Communication Group: Mariken van der Velden (coordinator), Loes Aaldering, Wouter van Atteveldt, Andreu Casas Salleras, Felicia Loecherbach, Anita van Hoof, Dirck de Kleer, Jan Kleinnijenhuis, Marloes Jansen, Nicolas Matthis, Kasper Welbers and Nienke Wolfers"
output: pdf_document
#  github_document:
#    toc: yes
editor_options: 
  chunk_output_type: console
always_allow_html: true
---


```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
#library(printr)
rm(list = ls())
```

# Scripts
- [Required Packages &amp; Reproducibility](#required-packages-&amp;-reproducibility)
- [Load Data](#load-data)
- [Recode Block Background Variables](#black-background-variables)
- [Recode Block A Voting Behavior](#recode-block-a-voting-behavior)
- [Recode Block B Performance Politics in the Media](#recode-block-b-performance-politics-in-the-media)
- [Recode Block H Other Issues](#recode-block-h-other-issues)
- [Recode Block I News Consumption](#recode-block-i-news-consumption)
- [Merge &amp; Save Data](#merge-&amp;-save-data)

## Required Packages &amp; Reproducibility
```{r, message=FALSE, warning=F}
rm(list=ls())

renv::snapshot()

library(tidyverse)
library(sjlabelled)
```

## Load Data
Load the data downloaded from Qualtrics, and only keep those that have given consent to participate in the panel study.

```{r}
f <- "../../data/raw-private/qualtrics_raw_wave1.csv"
col_names <- names(read_csv(f, n_max = 0))
d <- read_csv(f, col_names = col_names, skip = 3) %>%
  remove_all_labels() %>% 
  tibble() %>%
  #only keep people that have given consent
  filter(consent1==1 & consent2==1) 
rm(f, col_names)
```


## Recode Block Background Variables

```{r}
BG <- d %>%
  mutate(wave = "wave 1",
         start_date = StartDate,
         end_date = EndDate,
         duration_min = round(Duration..in.seconds./60,2)) %>%
  select(iisID, wave, start_date, end_date, duration_min, 
         progress = Progress)
```

## Recode Block A Voting Behavior
```{r}
A <- d %>%
  select(iisID, A1:A2, A2_otherparty = A2_14_TEXT,
         A2_DO_1:A2_DO_13, A3_DO_1:A3_DO_13) %>%
  mutate(A1 = recode(A1,
                     `2` = 0,
                     `3` = 998,
                     `4` = 999),
         A1_not = replace_na(A1_not, ""),
         A1_vote = replace_na(A1_vote, ""),
         A1_elab = paste0(A1_not, A1_vote),
         A2 = recode(A2,
                     `16` = 999),
         order_A2 = paste(A2_DO_1, A2_DO_2, A2_DO_3, A2_DO_4,
                          A2_DO_5, A2_DO_6, A2_DO_7, A2_DO_8,
                          A2_DO_9, A2_DO_10, A2_DO_11, 
                          A2_DO_12, A2_DO_13, sep = "|"),
         order_A3 = paste(A3_DO_1, A3_DO_2, A3_DO_3, A3_DO_4,
                          A3_DO_5, A3_DO_6, A3_DO_7, A3_DO_8,
                          A3_DO_9, A3_DO_10, A3_DO_11, 
                          A3_DO_12, A3_DO_13, sep = "|")) %>%
  select(iisID, A1, A1_elab, A2, A2_otherparty, order_A2, order_A3)
  
A3 <- d %>%
  select(iisID, A3_1:A3_13) %>%
  pivot_longer(cols = A3_1:A3_13,
               names_to = "variable") %>%
  drop_na(value) %>%
  separate(variable, c("variable", "party"), "_", extra = "merge")  %>%
  group_by(iisID) %>%
  summarise(n = row_number(),
            party = party) %>%
  ungroup() %>%
  mutate(n = paste("A3", n, sep="_"),
         n = factor(n),
         party = as.integer(party))

A3 <-  pivot_wider(A3, names_from = n, values_from = party, 
                    values_fill = 0) %>%
  mutate(A3 = paste(A3_1, A3_2, A3_3, A3_4, A3_5,
                    A3_6, A3_7, A3_8, A3_9, A3_10, 
                    A3_11, A3_12, sep = "|"))
A <- left_join(A, A3, by = "iisID") %>%
  select(iisID, A1:A2_otherparty, order_A2, A3, order_A3, A3_1:A3_12)
rm(A3)
```

## Recode Block B Performance Politics in the Media
```{r}
d <-d %>%
mutate(B2_1 = recode(B2_1,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_2 = recode(B2_2,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_3 = recode(B2_3,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_4 = recode(B2_4,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_5 = recode(B2_5,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_6 = recode(B2_6,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_7 = recode(B2_7,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_8 = recode(B2_8,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_9 = recode(B2_9,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_10 = recode(B2_10,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_11 = recode(B2_11,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_12 = recode(B2_12,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       B2_13 = recode(B2_13,`1` = 999, `2` = 1, `24` = 2,`25` = 3),
       order_B2 = paste(B2_DO_1, B2_DO_2, B2_DO_3, B2_DO_4, 
                          B2_DO_5, B2_DO_6, B2_DO_7, B2_DO_8, 
                          B2_DO_9, B2_DO_10, B2_DO_11, B2_DO_12, 
                          B2_DO_13, sep = "|"))

B3 <- d %>%
  select(iisID, X1_B3_3, X2_B3_3, X3_B3_3,
         X4_B3_3, X5_B3_3, X6_B3_3, X7_B3_3,
         X8_B3_3, X9_B3_3, X10_B3_3, X11_B3_3,
         X12_B3_3, X13_B3_3,
         X1_B3_4, X2_B3_4, X3_B3_4,
         X4_B3_4, X5_B3_4, X6_B3_4, X7_B3_4,
         X8_B3_4, X9_B3_4, X10_B3_4, X11_B3_4,
         X12_B3_4, X13_B3_4) %>%
  pivot_longer(cols = X1_B3_3:X13_B3_4,
               names_to = "variable") %>%
  separate(variable, c("variable", "question"), "_", extra = "merge")  %>%
  group_by(variable) %>%
  mutate(row = row_number()) %>%
  pivot_wider(names_from = question, values_from = value) %>%
  unite("B3", B3_3:B3_4, remove = T, na.rm = T) %>%
  mutate(B3 = ifelse(B3 == "", "999", B3),
         B3 = as.numeric(B3)) 
  
B3 <- B3 %>%
  group_by(variable) %>%
  mutate(row = row_number()) %>%
  pivot_wider(names_from = variable, values_from = B3) %>%
  select(iisID, B3_1 = X1,
         B3_2 = X1, B3_3 = X3, B3_4 = X4, B3_5 = X5,
         B3_6 = X6, B3_7 = X7, B3_8 = X1, B3_9 = X9,
         B3_10 = X10, B3_11 = X11, B3_12 = X12, B3_13 = X13) %>%
  distinct(iisID, .keep_all = TRUE)

B <- d %>%
  select(iisID, B2_1:B2_13, order_B2)
B <- left_join(B, B3, by = "iisID")
rm(B3)
```

# Recode Block H Other Issues
```{r}
H <- d %>%
  select(iisID, H1, H2, H7, H7a, H7b:H10) %>%
  mutate(H7 = recode(H7, `3` = 2, `4` = 3, `5` = 4)) 
```

# Recode Block I News Consumption
```{r}
I <- d %>%
  select(iisID, I1_1:I1_9, I2_1:I2_13, I2_other = I2_13_TEXT,
         I3_1:I3_14, I3_15 = I3_17,I3_16 = I3_15, I3_other = I3_15_TEXT,
         I4_1:I4_8, I4_9 = I4_15, I4_other_blogs = I4_8_TEXT,
         I4_other_sites = I4_15_TEXT,
         I5_1:I5_11, I5_other = I5_11_TEXT,
         I6_1:I6_6, I6_other = I6_6_TEXT,
         I7_1:I7_9, I7_10 = I7_11, I7_other = I7_11_TEXT) 

```

## Merge &amp; Save Data
```{r}
df <- left_join(BG, A, by = "iisID")
df <- left_join(df, B, by = "iisID")
df <- left_join(df, H, by = "iisID")
df <- left_join(df, I, by = "iisID") %>%
  distinct(iisID, .keep_all = T) %>%
  drop_na(A1) %>%
  filter(iisID !="007",
         iisID !="009") %>%
  mutate(iisID = as_numeric(iisID)) %>%
  drop_na(iisID)

write_csv(df, "../../data/intermediate/wave1.csv")
rm(A, B, H, I)
```

