---
title: 'VU 2021 Election Study: Respondents per media use categories'
author: "Wouter van Atteveldt, Nel Ruigrok, Mariken van der Velden"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo=F, fig.path='figures/', fig.width = 10)
library(printr)
options(digits = 2)
```

```{r}
#remotes::install_github("dill/emoGG")
library("emoGG")
library(tidyverse)
library(here)
source(here("src/lib/create_labels.R"))
data = list()
d <- read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv"))  %>% 
  select(iisID, 
         stemintentie = A2, 
         Gender = gender, 
         Leeftijd = age, 
         Onderwijsniveau = education, 
         Regio = region, 
         Etniciteit = ethnicity, 
         Werk = job, 
         Internetgebruik = internet_use, 
         vorige_stem=E1,
         covid_maatregelen=H3, 
         covid_vaccinatie=H7,
         pol_kennis=D1,
         linksrechts=E2,
         eu = H4
  ) %>%
  mutate(stemintentie = recode_parties(label(stemintentie, "A2")),
         Werk = label(Werk, "job"),
         Leeftijd = label(Leeftijd, "age"),
         Onderwijsniveau = label(Onderwijsniveau, "education"),
         Gender = label(Gender, "gender"),
         Regio = label(Regio, "region"),
         Internetgebruik = label(Internetgebruik, "internet_use"),
         Etniciteit = label(Etniciteit, "ethnicity"),
         vorige_stem = recode_parties(label(vorige_stem, "E1")),
         covid_maatregelen=label(covid_maatregelen, "H3"),
         covid_vaccinatie=label(covid_vaccinatie, "H7"))

m = long_mc() %>% filter(question =="I1") %>% select(-question) %>% pivot_wider(names_from = "medium") %>% 
  rename(tv=Television,
    newspaper=`Newspapers or opinion magazines (paper or online)`, 
             radio=`Radio (including podcast and online)`, 
             online=`Online news sites (such as nu.nl) or blogs`, 
             social=`Social media (such as Facebook, Twitter)`, 
             chat=`Messaging apps (such as Whatsapp, messenger)`,
             interpersonal=`(Offline) conversations with people`,
             apps=`News apps or push messages on your phone`,
             search=`Search engines (such as Google or Bing)`)

```


# Use types

We categorize users according to their most used source, 
normalized with respect to how often each source is used on average. 
Users that use mostly social media, messaging apps or search engines to find news are categorized as 'Social Media Users' (ðŸ¦). Users that mostly rely on (print) newspapers or radio are classified as 'Traditional Media Users' (ðŸ“°),
while users that mostly use online news sites or news apps are classified as 'online media users' (ðŸ“±).
Users that mostly use television are classified as TV users (ðŸ“º). 
Finally, users that use each media type less than average are seen as news avoiders (ðŸš«).

```{r}

x = m %>% select(iisID, tv:search) %>% pivot_longer(-iisID, names_to="medium") %>% 
  filter(medium %in% c("newspaper", "social", "apps", "online", "tv")) %>% 
  group_by(medium) %>% mutate(value=(value-mean(value))/sd(value)) %>% 
  arrange(iisID, -value) %>% group_by(iisID) %>%filter(row_number()==1) 

x = x %>% mutate(cat2 = case_when(value<0 ~ "none",
                              medium %in% c("social", "chat", "search") ~ "social",
                              medium %in% c("newspaper", "radio") ~ "newspaper",
                              medium %in% c("online", "app") ~ "online",
                              medium == "tv" ~ "tv"))

m2 = x %>% select(iisID, cat=cat2)
m2 %>% group_by(cat) %>% na.omit() %>% summarize(perc=n()) %>% mutate(perc=perc/sum(perc))
```

# Demographics per user type

```{r}

x = m2 %>% select(iisID, cat) %>% inner_join(d) %>% select(iisID, cat, Gender, Leeftijd, Onderwijsniveau, pol_kennis, linksrechts) %>%
  mutate(Female = as.numeric(Gender)-1, Gender=NULL,
         Leeftijd=as.numeric(Leeftijd), 
         Onderwijsniveau=ifelse(Onderwijsniveau=="Don't know", NA, 3-as.numeric(Onderwijsniveau)) ) %>% 
  pivot_longer(Leeftijd:Female) %>%
  group_by(cat, name) %>% summarize(value=mean(scales::rescale(value), na.rm=T))

data$Demographics = x

ggplot(mapping=aes(x=name, y=value, emoji=cat)) + ylim(0.2,.8) +
  geom_emoji(data=filter(x, cat=="newspaper"), emoji="1f4f0") +
  geom_emoji(data=filter(x, cat=="tv"), emoji="1f4fa") +
  geom_emoji(data=filter(x, cat=="social"), emoji="1f426") +
  geom_emoji(data=filter(x, cat=="online"), emoji="1f4f1") +
  geom_emoji(data=filter(x, cat=="none"), emoji="1f6ab") +
  coord_flip() + ggthemes::theme_hc() + ggtitle("Demographics per media user type", "(social, online, newspaper, tv, none)")
```

# Vote intention per user type

```{r}
x = m2 %>% select(iisID, cat) %>% inner_join(d) %>% select(iisID, cat, stemintentie) %>% group_by(cat, stemintentie) %>% summarize(n=n()) %>% mutate(n=n/sum(n)) 

data$Stemintentie = x

ggplot(mapping=aes(x=stemintentie, y=n)) +
  geom_emoji(data=filter(x, cat=="newspaper"), emoji="1f4f0") +
  geom_emoji(data=filter(x, cat=="tv"), emoji="1f4fa") +
  geom_emoji(data=filter(x, cat=="social"), emoji="1f426") +
  geom_emoji(data=filter(x, cat=="online"), emoji="1f4f1") +
  geom_emoji(data=filter(x, cat=="none"), emoji="1f6ab") +
  coord_flip() + ggthemes::theme_hc() + ggtitle("Vote intention per media user type", "(percentage within user type)")
```


```{r write-data}
writexl::write_xlsx(data, here("src/analysis/categories.xlsx"))
```

