---
title: 'VU 2021 Election Study: Analysis of pre-wave'
author: "Wouter van Atteveldt, Mariken van der Velden, Nel Ruigrok"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.path='figures/')
#library(printr)
library(printr)
options(digits = 2)
library(here)
library(tidyverse)
source(here("src/lib/create_labels.R"))
```


# Data

```{r "Read Data"}
d = read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv"))
```

# Vote intention 

```{r "Prepare data"}
votes = d %>% select(iisID, A2, E1) %>% 
  mutate(A2=recode_parties(label(A2, "A2")), E1=recode_parties(label(E1, "E1")))
rep = votes %>% pivot_longer(A2:E1, names_to="when", values_to="party") %>% 
  group_by(party, when) %>% summarize(n=n()) %>% ungroup() %>% 
  pivot_wider(names_from="when", values_from="n") %>% 
  filter(!is.na(party), !party %in% c("Niet", "Nieuw", "Weet niet")) %>% 
  mutate(A2=A2/sum(A2)*150, E1=E1/sum(E1)*150) %>% 
  left_join(read_csv(here("data/raw/tk2017.csv"))) %>% replace_na(list(tk=0)) %>% 
  mutate(overrep=E1 - tk)
```

## Overrepresentation E1 wrt actual distribution TK2017

Note: A mismatch between self-reported votes and actual distribution
can be caused by unrepresentative sample (esp. since our data is unweighted)
and misremembering by respondents (e.g. accidentally remembering their provincial vote)

```{r}
rep %>% arrange(-abs(overrep)) %>% select(-A2)
```

## Overall changes in voting

```{r}
rep %>% mutate(change=A2-E1) %>% select(party, E1, A2, change) %>% arrange(-abs(change))
```

## How do voters move between parties?

```{r "wave0_move", fig.width=10, fig.height=10}
library(igraph)
gdf = votes %>% 
  
  filter(A2 != "Weet niet", E1 != "Weet niet") %>% 
  group_by(A2, E1) %>% summarize(n=n()) %>% na.omit() %>% filter(A2 != E1)

g = full_join(gdf, gdf, by=c("A2"="E1", "E1"="A2")) %>% replace_na(list("n.x"=0, "n.y"=0)) %>% mutate(n=n.x - n.y) %>% select(E1, A2, n) %>% filter(n>0) %>% 
  filter(n>5) %>% graph_from_data_frame()

E(g)$width = 1+E(g)$n/10 
E(g)$label = round((E(g)$n / 1783) * 150, 1)
E(g)$arrow.size=E(g)$width/2

plot(g)
```

# Media usage

## Data

```{r}
library(glue)
result = list()
for (var in str_c("I", 1:7)) 
  result[[var]] = d %>% select(iisID, starts_with(var), -contains("_other")) %>% pivot_longer(-iisID, names_to="medium") %>% 
    mutate(medium=label(as.numeric(str_remove(medium, glue("^{var}_"))), var)) %>% filter(!is.na(value))
newsuse = bind_rows(result, .id="question") %>% mutate(medium=str_remove_all(medium, " \\(.*"))

resp = d %>% select(iisID, age, education) %>% mutate(age=label(age, "age"), education=label(education, "education"))
overall = newsuse %>% filter(question=="I1") %>% inner_join(resp) %>% 
  mutate(age2=case_when(age %in% c("<24" ,"25-34") ~ "<35",
                        age %in% c("55-64", ">64") ~ "55+",
                        T ~ "35-55"))
```

## Overall use of various media per age/education

```{r "wave0_mediause"}
news_agg = overall %>% group_by(age2, education, medium) %>% summarize(value=mean(value), n=n())
news_agg %>% filter(education!="Don't know") %>% mutate(medium=fct_reorder(medium, value)) %>% 
  ggplot(aes(x=education, y=age2, label=round(value, 1), fill=value)) + 
  geom_tile()+ geom_text()+
  facet_wrap("medium") + 
  ggtitle("Overall use of different media per age and education group", "(in days per week)")  + 
  xlab("Education group (high=University/HBO, med=HAVO/MBO2+, low=MAVO/MBO1") + ylab("Age group") +
  scale_fill_gradient(low="white", high="#3182bd", guide=F)
```

