---
title: 'VU 2021 Election Study: Issues of COVID-19 and Childs Benefits Scandal in Dutch Elections'
author: "Mariken van der Velden"
permalink: reports/deutsche-welle/
output: 
  md_document:
    toc: yes
    preserve_yaml: yes
---

```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
rm(list = ls())
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      echo = FALSE, fig.align = "center",
                        fig.path='')
library(printr)
options(digits = 2)
library(here)
library(tidyverse)
library(viridis)
source(here("src/lib/data.R"))
```

```{r "data"}
bg <- read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv")) %>%
  select(iisID, age, gender, education, region, G1_b)
d <- read_csv(here("data/intermediate/wave2.csv")) %>%
  select(iisID, H_corona_1:H_corona_11, H_corona2, A2, B2_12, B3_12)

d <- left_join(d, bg, by = "iisID") %>%
  mutate(A2 = recode(A2, 
                     `1` = "CDA",
                     `2` = "CU",
                     `3` = "D66",
                     `4` = "Denk",
                     `5` = "FvD",
                     `6` = "GL",
                     `7` = "PvdA",
                     `8` = "PvDD",
                     `9` = "PVV",
                     `10` = "SGP",
                     `11` = "SP",
                     `12` = "VVD",
                     `13` = "50+",
                     `14` = "Don't Know", 
                     `15` = "Don't Know",
                     `999` = "Don't Know"),
         H_corona2 = recode(H_corona2,
                            `6` = 2,
                            `7` = 3,
                            `8` = 4,
                            `9` = 5)) 
 
```
# Child Benefits Scandal in the Dutch Elections
On average 33% of the respondents thought that the government had to step down over the child benefits scandal (visualized by the dashed line in the figure below).
If you intend to vote for VVD or CDA (two of the current (interim) government parties), you are more likely to say that the government should not have fallen over this.
If you intent to vote for the smaller (interim) government parties, CU or D66, you are more likely to say that the government was right to step down of the child benefits scandal.
Other than current interim government parties, people intending to vote for the animal rights party (PvdD), the greens (GL), or Baudet's party (FvD) were less in favor of the government stepping down. 
If you intend to vote for the socialist party (SP), Denk, the orthodox christian party (SGP), or the elderly party (50+), you are more likely to say that the government was right to step down of the child benefits scandal.

```{r "Child-Benefits-Party", fig.height=10, fig.width=10}
df <- d %>%
  mutate(G1_b = ifelse(G1_b < 4, "Yes: Government had to step down",
                ifelse(G1_b > 3 & G1_b <7, "Neutral",
                ifelse(G1_b > 6 & G1_b < 11, "No: Government shouldn't have fallen",
                G1_b))),
         G1_b = factor(G1_b, 
                       levels = c("No: Government shouldn't have fallen",
                                  "Neutral", 
                                  "Yes: Government had to step down"))) %>%
  drop_na(A2, G1_b) %>%
  group_by(A2, G1_b) %>%
  summarise(n = n()) %>%
  mutate(percent = round(n/sum(n),2))

ggplot(df, aes(y = A2, x = percent, color = G1_b)) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = .33, size = .3, linetype = "dashed") +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "", 
       x = " Government had to step down (0) - Government shouldn't have fallen (10)",
       title = "Childs Benefits Scandal: Is the Government to Blame?",
       subtitle = "Mean per Party Preference") +
  scale_color_viridis_d() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position="bottom",
        legend.title = element_blank()) +
  guides(color=guide_legend(nrow=1,byrow=TRUE))
```

Download data: [`r export_data(df, "Figure Childs Benefits Scandal - Party")`]

Looking at how the Child Benefits Scandal affects the evaluation of Rutte, we see that -- as expected -- if you positively evaluate Rutte, you are more likely to say that the government shouldn't have fallen and vice versa.

```{r "Child-Benefits-Rutte", fig.height=10, fig.width=10}
df <- d %>%
  mutate(G1_b = ifelse(G1_b < 4, "Yes: Government had to step down",
                ifelse(G1_b > 3 & G1_b <7, "Neutral",
                ifelse(G1_b > 6 & G1_b < 11, "No: Government shouldn't have fallen",
                G1_b))),
         G1_b = factor(G1_b, 
                       levels = c("No: Government shouldn't have fallen",
                                  "Neutral", 
                                  "Yes: Government had to step down")),
         B3_12 = ifelse(B3_12 < 4, "Negative Evaluation of Rutte",
                 ifelse(B3_12 > 3 & B3_12 <7, "Neutral Evaluation of Rutte",
                 ifelse(B3_12 > 6 & B3_12 < 11, "Positive Evaluation of Rutte",
                 ifelse(B3_12 == 999, NA, B3_12))))) %>%
  drop_na(B3_12, G1_b) %>%
  group_by(B3_12, G1_b) %>%
  summarise(n = n()) %>%
  mutate(percent = round(n/sum(n),2)) 

ggplot(df, aes(y = G1_b, x = percent, color = B3_12)) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = .33, size = .3, linetype = "dashed") +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "", 
       x = " Government had to step down (0) - Government shouldn't have fallen (10)",
       title = "Childs Benefits Scandal: Is the Government to Blame?",
       subtitle = "Mean per group") +
  scale_color_viridis_d() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position="bottom",
        legend.title = element_blank()) +
  guides(color=guide_legend(nrow=1,byrow=TRUE))
```

Download data: [`r export_data(df, "Figure Childs Benefits Scandal - Rutte")`]

# Covid-19 in the Dutch Elections

The most important issue in the elections of 2021 is Covid-19.
Accross the board, people do support the Corona measures (average of 7.3 on a scale from 0-10), visualized by the dashed line in the figure below.
The figure below also shows some variations over parties and between measures.
On average, there's least support for the curfew and for opening shops, bars, and restaurants accross all parties.
Moreover, between parties, if you intend to vote Baudet's party (FvD), Wilders' party (PVV) or Denk, you're least supportive of the Corona measures.
If you intend to vote for any of the other parties, you are quite supportive of the measures, with unsurpisingly people voting for the elderly party (50+) to be most supportive.

```{r "Support-Corona-Measures-Party", fig.height=10, fig.width=10, fig.cap="Support Corona Measures"}
df <- d %>%
   drop_na(A2) %>%
  group_by(A2) %>%
  summarise(`Work from home` = mean(H_corona_1, na.rm = T),
            `Washing hands` = mean(H_corona_2, na.rm = T),
            `Social distancing` = mean(H_corona_3, na.rm = T),
            `Wearing Mask` = mean(H_corona_6, na.rm = T),
            `Testing if symptomatic` = mean(H_corona_7, na.rm = T),
            `Staying home after testing yourself or anyone in household`  = mean(H_corona_8, na.rm = T),
            `Quarantaine after traveling` = mean(H_corona_9, na.rm = T),
            `Curfew` = mean(H_corona_10, na.rm = T),
            `Shops, bars, and restaurants open` = mean(H_corona_11, na.rm = T)) %>%
  pivot_longer(cols = `Work from home`:`Shops, bars, and restaurants open`) %>%
  mutate(means = mean(value, na.rm = T))

ggplot(df, aes(y = A2, x = value, color = name)) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = 7.3, size = .3, linetype = "dashed") +
  labs(y = "", 
       x = "Fully disagree (0) - Fully agree (10)",
       title = "Support for Corona Measures",
       subtitle = "Mean per Party Preference") +
  scale_color_viridis_d() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position="bottom",
        legend.title = element_blank()) +
  guides(color=guide_legend(nrow=4,byrow=TRUE))
```

Download data: [`r export_data(df, "Figure Support Corona Measures - Party")`]

Looking at the evaluation for PM Rutte based on the Corona measures (see figure below), we see that there are hardly differences between your support for the Corona measures and how well you evaluate Rutte, except for the measures _Curfew_, _Opening of Shops, Bars and Restaurants_, and _Wearing Masks_. 

```{r "Support-Corona-Measures-Rutte", fig.height=10, fig.width=10, fig.cap="Support Corona Measures"}
df <- d %>%
  mutate(B3_12 = ifelse(B3_12 < 4, "Negative Evaluation of Rutte",
                 ifelse(B3_12 > 3 & B3_12 <7, "Neutral Evaluation of Rutte",
                 ifelse(B3_12 > 6 & B3_12 < 11, "Positive Evaluation of Rutte",
                 ifelse(B3_12 == 999, NA, B3_12))))) %>%
  drop_na() %>%
  group_by(B3_12) %>%
  summarise(`Work from home` = mean(H_corona_1, na.rm = T),
            `Washing hands` = mean(H_corona_2, na.rm = T),
            `Social distancing` = mean(H_corona_3, na.rm = T),
            `Wearing Mask` = mean(H_corona_6, na.rm = T),
            `Testing if symptomatic` = mean(H_corona_7, na.rm = T),
            `Staying home after testing yourself or anyone in household`  = mean(H_corona_8, na.rm = T),
            `Quarantaine after traveling` = mean(H_corona_9, na.rm = T),
            `Curfew` = mean(H_corona_10, na.rm = T),
            `Shops, bars, and restaurants open` = mean(H_corona_11, na.rm = T)) %>%
  pivot_longer(cols = `Work from home`:`Shops, bars, and restaurants open`) %>%
  mutate(means = mean(value, na.rm = T))

ggplot(df, aes(y = name, x = value, color = B3_12)) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = 7.3, size = .3, linetype = "dashed") +
  labs(y = "", 
       x = "Fully disagree (0) - Fully agree (10)",
       title = "Support for Corona Measures",
       subtitle = "Mean per group") +
  scale_color_viridis_d() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position="bottom",
        legend.title = element_blank()) +
  guides(color=guide_legend(nrow=3,byrow=TRUE))
```

Download data: [`r export_data(df, "Figure Support Corona Measures - Rutte")`]

```{r, eval = F}
d %>%
  group_by(A2) %>%
  summarise(aan_houden = mean(H_corona2, na.rm = T))
```
