---
title: 'Media Visibility and Electoral shifts'
author: "Wouter van Atteveldt, Mariken van der Velden, Andreu Casas"
permalink: reports/wave2/
output: 
  md_document:
    preserve_yaml: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo=F, fig.path='', fig.width = 10)
library(printr)
options(digits = 2)
```

```{r data}
library(here)
library(tidyverse)
library(amcatr)

source(here("src/lib/data.R"))
source(here("src/lib/amcatlib.R"))
data = list(w0 = read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv")),
            w1 = read_csv(here("data/intermediate/wave1.csv")),
            w2 = read_csv(here("data/intermediate/wave2.csv"))
)

resp = extract_wide(data$w0)
w = extract_waves(data)
votes = bind_rows(
  w %>% filter(variable == "vote") %>% select(wave, iisID, name),
  #resp %>% select(iisID, vote_2017) %>% rename(name = vote_2017) %>% add_column(wave="k17")
)
wave_label= function(wave) case_when(wave == "w0" ~ "January", wave == "w1" ~ "Early February", wave =="w2" ~ "Late February")

theme_nrc = function(...) {
  ggthemes::theme_hc(...) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          axis.title.x=element_blank(), 
          axis.title.y=element_blank(),
          strip.background = element_rect(fill="white"))
}
```

# Visibility of parties

```{r media-data}
period = function(date) {
  case_when(date < "2021-01-13" ~ "pre",
          date < "2021-01-28" ~ "w0",
          date < "2021-02-16" ~ "w1",
         # date < "2021-02-28" ~ "w2",
          T ~ "w2")
}
queries = get_party_queries()

tmp = here("data/tmp/wave2_meta.rmd")
if (file.exists(tmp)) {
  meta = read_rds(tmp)
} else {
  conn = amcat.connect("https://vu.amcat.nl")
  meta_print = amcat.articles(conn, project=69, articleset=2562, columns = c("date", "publisher")) %>%
    as_tibble() %>%  add_column(mtype="print")
  meta_online = amcat.articles(conn, project=69, articleset=2564, columns = c("date", "publisher")) %>%
    as_tibble() %>%  add_column(mtype="online") %>% filter(publisher != "Nieuwsuur") %>% mutate(publisher="nos.nl")
  meta_tv = amcat.articles(conn, project=69, articleset=2579, columns = c("date", "publisher")) %>% 
    as_tibble() %>% add_column(mtype="tv")
  meta = bind_rows(meta_print, meta_online, meta_tv) %>% mutate(id=as.numeric(id)) %>% mutate(period=period(date))
  write_rds(meta, tmp)
}

tmp = here("data/tmp/wave2_media_visibility.rmd")
if (file.exists(tmp)) {
  d = read_rds(tmp)
} else {
  conn = amcat.connect("https://vu.amcat.nl")
  h = amcat.hits(conn, project=69, sets=c(2562, 2564, 2579), queries=queries$query, labels = queries$partij) %>% 
    as_tibble() %>% rename(party=query)
  d = left_join(h, meta)
  write_rds(d, tmp)
}

totals = meta %>% group_by(period, mtype) %>% summarize(ntot=n())
parties = d %>% group_by(party) %>% summarize(n=n()) %>% arrange(-n) %>% pull(party)
```

```{r visibility-newspapers}
hpp = d %>% group_by(period, party, mtype) %>% summarize(n=n(), .groups="drop") %>% 
  pivot_wider(names_from = "party", values_from="n", values_fill=0) %>% pivot_longer(-(period:mtype), names_to="party", values_to="n") %>%
  inner_join(totals) %>%
  mutate(perc=n/ntot, period = as.numeric(as.factor(period)), party=factor(party, levels=parties))

hpp %>% filter(mtype == "print") %>% 
ggplot(aes(x=period, y=perc, color=party)) + geom_line() + 
  facet_wrap("party") + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Visibility of parties per wave: Newspapers", "(as percentage of all political news)") +
  scale_x_continuous(breaks=1:4, labels=c("1 Jan", "15 Jan", "1 Feb", "15 Feb")) + theme_nrc() +
  theme(legend.position = "none")
```

No large shifts in visibility in the newspapers. VVD and coalition parties VDA and D66 are most visible. Groenlinks, SP, and FvD are increasing in visibility,
while PvdA remains mostly constant. 

```{r visibility-tv}
hpp %>% filter(mtype == "tv") %>% 
ggplot(aes(x=period, y=perc, color=party)) + geom_line() + 
  facet_wrap("party") + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Visibility of parties per wave: NPO", "(as percentage of all political news)") +
  scale_x_continuous(breaks=1:4, labels=c("1 Jan", "15 Jan", "1 Feb", "15 Feb")) + theme_nrc() +
  theme(legend.position = "none")
```

On television (subtitles of NPO news and current affairs shows) we see the same overall pattern, but while D66 and VVD make gains, PvdA, GroenLinks, SP and CU all lose visibility.
In contrast, PVV and especially FvD gain visibility. 

```{r visibility-nosnl}
hpp %>% filter(mtype == "online") %>% 
ggplot(aes(x=period, y=perc, color=party)) + geom_line() + 
  facet_wrap("party") + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Visibility of parties per wave: NOS.nl (inc. liveblogs)", "(as percentage of all political news)") +
  scale_x_continuous(breaks=1:4, labels=c("1 Jan", "15 Jan", "1 Feb", "15 Feb")) + theme_nrc() +
  theme(legend.position = "none")

```

Here we do see a clear shift, with D66 even overtaking VVD and CDA, while the other leftist parties lose some of their gains. 
PVV and the Smaller parties get more visibility, but GL, and FvD actually become less visible. 

## Visibility and engagement on social media

```{r visibility-sm}
sm = read_csv(here("data/raw/messages-shares-likes-mentions-all-leaders-1jan2021-03march2021.csv"))%>% 
  mutate(party=str_match(name, "\\((.*)\\)")[,2],
              variable=str_remove(variable, "Number of ")) %>% 
  select(platform, variable, party, date=day_only, value)
sma = sm %>% mutate(period=period(date),
              variable2=ifelse(variable=="Messages", "messages", "engagement")
              ) %>% group_by(period, variable2, party) %>% summarize(value=sum(value)) %>% mutate(perc=value/sum(value))
sma %>% filter(variable2=="messages") %>% 
  ggplot(aes(x=as.numeric(as.factor(period)), y=perc, color=party)) + geom_line() + facet_wrap("party")+ 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Social Media Activity of parties per wave", "(# of messages on FB, IG, TW as percentage of all messages)") +
  scale_x_continuous(breaks=1:4, labels=c("1 Jan", "15 Jan", "1 Feb", "15 Feb")) + theme_nrc() +
  theme(legend.position = "none")
```

```{r engagement-sm}
sma %>% filter(variable2=="engagement") %>% 
  ggplot(aes(x=as.numeric(as.factor(period)), y=perc, color=party)) + geom_line() + facet_wrap("party")+ 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Social Media Engagement with parties per wave", "(Shares, likes, mentions of FB, IG, TW posts as percentage of all shares/likes/mentions)") +
  scale_x_continuous(breaks=1:4, labels=c("1 Jan", "15 Jan", "1 Feb", "15 Feb")) + theme_nrc() +
  theme(legend.position = "none")
```

These two graphs show a very clear picture: although all parties are active on social media, FvD wins hands down in terms of their activity,
and FvD and PVV are the only parties with substantial engagement on social media (likes, shares, mentions)

# Does the voter take notice?

[TODO]

# The Undecided Voter

The number of undecided voters (and non-voters) is becoming smaller as the election draws near:

```{r undecided0}
w0u = votes %>% filter(name == "Undecided", wave == "w0") %>% pull(iisID)
w1u = votes %>% filter(name == "Undecided", wave == "w1") %>% pull(iisID)
w2u = votes %>% filter(name == "Undecided", wave == "w2") %>% pull(iisID)

votes  %>% group_by(wave, name) %>% summarize(n=n())%>% mutate(wave=wave_label(wave), perc=n/sum(n)*100) %>% filter(name %in% c("Undecided", "Not")) %>% select(-n) %>% pivot_wider(names_from=wave, values_from=perc)

```

## Media use of undecided voters

Last month, we saw that undecided voters were on average less interested in news and politics. 
If we look at the group of voters that were undecided at the start of the campaign, 
how did their media use change?

```{r undecided-media}
d = w %>% mutate(zwevend=fct_rev(ifelse(iisID %in% w0u, "Undecided", "Decided"))) %>% filter(variable == "I1") %>% 
  mutate(name=case_when(name == "Television" ~ "TV",
                        name == "Newspapers or opinion magazines (paper or online)" ~ "Newspapers",
                        name == "Online news sites (such as nu.nl) or blogs" ~ "Online",
                        name == "Social media (such as Facebook, Twitter)" ~ "Social",
                        T ~ "DROP")) %>% 
  filter(name != "DROP") %>% group_by(zwevend, wave, name) %>% summarize(value=mean(value)) 
ggplot(d, aes(x=as.numeric(as.factor(wave)), colour=name, y=value)) + geom_line() + facet_grid(cols=vars(zwevend)) + 
  theme_nrc()  + ggtitle("Media use of (un)decided voters", "(in frequency of use per week, 0-7)") +
  scale_x_continuous(breaks=c(1,2,3), labels=c("Wave 1\n(Late Jan.)", "Wave 2\n(Early Feb.)", "Wave 3\n(Late Feb.)"))

```

[Download data: `r export_data(d, "Media use of (un)decided voters")`]

These results are quite striking: much more than the voters that had already made up their mind,
these undecided voters started consuming more media, both newspapers and TV,
and at the end of February even consume more news. 

## Who are the undecided voters?

As undecided voters make up their mind, and other voters start doubting their choices, the group of undecided voters changes.
Let's look at some of the demographics:

### Education of undecided voters per wave

```{r undecided-education}
undecideds = w %>% filter(variable == "vote", name=="Undecided") %>% select(wave, iisID) %>% inner_join(resp)
undecideds %>% group_by(wave, education) %>% summarize(n=n()) %>% na.omit() %>% mutate(perc=n/sum(n)*100) %>% select(-n) %>% pivot_wider(names_from=wave, values_from=perc)
```

A small shift from medium to lower education. 

### Age of undecided voters per wave

```{r undecided-age}
undecideds %>% group_by(wave, age) %>% summarize(n=n()) %>% na.omit() %>% mutate(perc=n/sum(n)*100) %>% select(-n) %>% pivot_wider(names_from=wave, values_from=perc)
```

Here is a clear shift: where the undecided voters at the start of the campaign were relatively young,
by late February the group is older, with almost a quarter being 65 or older
and much fewer being younger than 25. 


### Other demographics of undecided voters per wave

```{r undecided-demo}
undecideds %>% mutate(education=2-as.numeric(education), age=as.numeric(age)) %>% 
   group_by(wave) %>% summarize(rile_self=mean(rile_self, na.rm=T), polknow=mean(polknow, na.rm=T), education=mean(education, na.rm=T), age=mean(age, na.rm=T)) %>%
  pivot_longer(-wave, names_to="Measure") %>% pivot_wider(names_from=wave)
```

The top rows shows the self-placement on a left-right scale, which shows that the group of undecided voters has shifted mostly towards the mean of left-right placement of 5.4. 
Interestingly, even though the group has a lower overall education, they score better on political knowledge, probably as a result of being older. 

(The bottom rows show education and age as a numeric value, mirroring the shift towards an older and slightly less educated group of voters. )

## Where do undecided voters go to?

```{r undecideds}
parties = c("Undecided", "VVD", "PVV", "CDA", "D66", "GroenLinks", "SP", "PvdA", "ChristenUnie", "PvdD", "50+", "SGP", "Denk", "FvD", "Other", "Not")

#votes %>% filter(name == "Undecided")
w1u = votes %>% filter(name == "Undecided", wave == "w1") %>% pull(iisID)
w2u = votes %>% filter(name == "Undecided", wave == "w2") %>% pull(iisID)
perc_w2 = length(intersect(w1u, w2u)) / length(w1u)

votes = votes %>% na.omit() %>% mutate(waveno=as.numeric(as.factor(wave)), nxt= waveno+1)# %>% mutate(name=ifelse(name %in% c("Not", "Undecided"), "Not/Undecided", name)) 
changes = votes %>% inner_join(votes, by=c("iisID", "nxt"="waveno"), suffix=c("", ".to")) %>% rename(to=name.to) %>% 
  group_by(wave, wave.to, name, to) %>% summarize(n=n())%>% ungroup()

undecided_to = changes %>% filter(wave != "k17", name %in%  c("Not", "Undecided")) %>% group_by(wave, to) %>% summarize(n=sum(n)) #%>% mutate(perc=n/sum(n)*100)%>% select(-n) %>% 
  #pivot_wider(values_fill=0, names_from=wave, values_from=perc) %>% pivot_longer(-to, values_to="perc", names_to="wave")

undecided_from = changes %>% filter(wave != "k17", to %in%  c("Not", "Undecided")) %>% group_by(wave,name) %>% summarize(n=sum(n)) #%>% #mutate(perc=n/sum(n)*100)%>% select(-n) %>% 
  #pivot_wider(values_fill=0, names_from=wave, values_from=perc) %>% pivot_longer(-name, values_to="perc", names_to="wave")

totals = votes %>% group_by(wave) %>% summarize(ntot=n())

d = bind_rows(
  undecided_to %>% rename(party=to) %>% add_column(type="Undecided voters went to"),
  undecided_from %>% rename(party=name) %>% add_column(type="Undecided came from"),
  tibble(wave="w0", party="Denk", n=0, type="Undecided came from"))  %>% 
  mutate(period=case_when(wave == "w0" ~ "Early February", T ~ "Late February"))

change = d %>% select(party, wave, period, type, n) %>% pivot_wider(names_from=type, values_from=n) %>% 
  left_join(totals) %>% 
  mutate(change=`Undecided voters went to` - `Undecided came from`,
         perc=change/ntot*100) %>% 
  select(party, wave, period, n=change, perc) %>% add_column(type="Net Change")


d2 = bind_rows(d, change) 
d2 = d2 %>% left_join(totals) %>% filter(party!="Undecided") %>% mutate(
  perc = n/ntot*150,
  colval=ifelse(type=="Net Change", perc*3, perc), 
  type=fct_rev(type),
  party2=fct_rev(factor(party, levels=parties)))

ggplot(d2, aes(x=period, y=party, fill=colval, label=round(perc, 1))) + geom_tile(color="white", lwd=2) + geom_text(color="white") +
  facet_grid(cols=vars(type)) +
  scale_fill_gradient2(low=scales::muted("red"), mid="#e0e0e0", high="#01529d", guide=F) + 
  xlab(NULL) + ylab(NULL) + ggtitle("Where do undecided voters go to and come from?", "(as virtual seats)") +
  ggthemes::theme_hc() + theme(panel.grid.major.y = element_blank(), strip.background = element_rect(fill="white"))
```

[Download data: `r export_data(d, "Undecided Voters")`]

On the left hand side, the figure above shows where undecided voters go to when they make up their mind.
Similarly, the middle columns show where new undecided voters go to. 
Finally, the rightmost two columns show the difference between the two: 
was a party able to attract more undecided voters they lost? 

The first thing that is very striking is that many party both attract a substantial number of undecided voters, while at the same time losing voters that become undecided. 
In fact, only `r perc_w2`% of undecided voters in wave 2 were also undecided in wave 1!
A lot of respondents switch in and out of the undecided group. 

Looking at which parties gain or lose from undecided voters, overall most parties make net gains as the number of undecided voters goes down.
VVD, Groenlinks and PvdA and CU are mostly able to profit in early February, with the latter three even making a net loss later that month.
Apparently, many voters started doubting their choice for these parties.
In contrast, CDA and D66 but also PvdD and FvD actually profited most in the second half of February. 

