---
title: 'Media Visibility and Electoral shifts'
author: "Wouter van Atteveldt, Mariken van der Velden, Andreu Casas"
permalink: reports/wave2/
output: 
  md_document:
    preserve_yaml: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo=F, fig.path='', fig.width = 10)
library(printr)
options(digits = 2)
```

```{r data}
library(here)
library(tidyverse)
library(amcatr)

source(here("src/lib/data.R"))
source(here("src/lib/amcatlib.R"))
data = list(w0 = read_csv(here("data/intermediate/VUElectionPanel2021_wave0.csv")),
            w1 = read_csv(here("data/intermediate/wave1.csv")),
            w2 = read_csv(here("data/intermediate/wave2.csv")),
            w3 = read_csv(here("data/intermediate/wave3.csv"))
)

resp = extract_wide(data$w0)
w = extract_waves(data)
votes = bind_rows(
  w %>% filter(variable == "vote") %>% select(wave, iisID, name),
  #resp %>% select(iisID, vote_2017) %>% rename(name = vote_2017) %>% add_column(wave="k17")
)

table(w$wave)
wave_label= function(wave) case_when(wave == "w0" ~ "January", wave == "w1" ~ "Early February", wave =="w2" ~ "Late February", "w3" ~ "Early March")

theme_nrc = function(...) {
  ggthemes::theme_hc(...) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          axis.title.x=element_blank(), 
          axis.title.y=element_blank(),
          strip.background = element_rect(fill="white"))
}
```

# Visibility of parties

```{r media-data}
period = function(date) {
  case_when(date < "2021-01-13" ~ "pre",
          date < "2021-02-01" ~ "w0",
          date < "2021-02-16" ~ "w1",
          date < "2021-03-01" ~ "w2",
          T ~ "w3")
}
#data$w0 %>% group_by(lubridate::floor_date(start_date, 'day')) %>% summarize(w0=n()) %>% 
#  full_join(data$w1 %>% group_by(lubridate::floor_date(start_date, 'day')) %>% summarize(w1=n())) %>% 
#  full_join(data$w2 %>% group_by(lubridate::floor_date(start_date, 'day')) %>% summarize(w2=n())) %>% 
#  full_join(data$w3 %>% group_by(lubridate::floor_date(start_date, 'day')) %>% summarize(w3=n())) %>%
#  View()

queries = get_party_queries()

tmp = here("data/tmp/wave3_meta.rmd")
if (file.exists(tmp)) {
  meta = read_rds(tmp)
} else {
  conn = amcat.connect("https://vu.amcat.nl")
  meta_print = amcat.articles(conn, project=69, articleset=2562, columns = c("date", "publisher")) %>%
    as_tibble() %>%  add_column(mtype="Newspapers")
  meta_online = amcat.articles(conn, project=69, articleset=2564, columns = c("date", "publisher")) %>%
    as_tibble() %>%  add_column(mtype="Online (nos.nl/nu.nl)") %>% filter(publisher %in% c("NOS nieuws", "NU.nl"))
  meta_tv = amcat.articles(conn, project=69, articleset=2579, columns = c("date", "publisher")) %>% 
    as_tibble() %>% add_column(mtype="TV")
  meta = bind_rows(meta_print, meta_online, meta_tv) %>% mutate(id=as.numeric(id)) %>% mutate(period=period(date))
  write_rds(meta, tmp)
}
tmp = here("data/tmp/wave2_media_visibility.rmd")
if (file.exists(tmp)) {
  d = read_rds(tmp)
} else {
  conn = amcat.connect("https://vu.amcat.nl")
  h = amcat.hits(conn, project=69, sets=c(2562, 2564, 2579), queries=queries$query, labels = queries$partij) %>% 
    as_tibble() %>% rename(party=query)
  d = left_join(h, meta)
  write_rds(d, tmp)
}

totals = meta %>% group_by(period, mtype) %>% summarize(ntot=n())
parties = d %>% group_by(party) %>% summarize(n=n()) %>% arrange(-n) %>% pull(party)
```

```{r visibility-newspapers}
hpp = d %>% group_by(period, party, mtype) %>% summarize(n=n(), .groups="drop") %>% 
  pivot_wider(names_from = "party", values_from="n", values_fill=0) %>% pivot_longer(-(period:mtype), names_to="party", values_to="n") %>%
  inner_join(totals) %>%
  mutate(perc=n/ntot, period = as.numeric(as.factor(period)), party=factor(party, levels=parties))

sm = read_csv(here("data/raw/messages-shares-likes-mentions-all-leaders-1jan2021-11march2021.csv"))%>% 
  mutate(party=str_match(name, "\\((.*)\\)")[,2],
              variable=str_remove(variable, "Number of ")) %>% 
  select(platform, variable, party, date=day_only, value)

sma = sm %>% mutate(period=period(date),
              mtype=ifelse(variable=="Messages", "Social Media (activity)", "Social Media (engagement)")
              ) %>%
  group_by(period, mtype, party) %>% summarize(n=sum(value)) %>% 
  mutate(perc=n/sum(n)) %>% ungroup() %>% 
  mutate(period=as.numeric(as.factor(period)), 
         party=case_when(party == "50PLUS" ~ "50Plus", party == "DENK" ~ "Denk", T~ party),
         party=factor(party, levels=parties))

hpp2=bind_rows(hpp, sma)
table(d$period)
mtypes = c("Newspapers"=1,"TV"=1, "Online (nos.nl/nu.nl)"=1, "Social Media (activity)"=2, "Social Media (engagement)"=2)
hpp2 %>% mutate(mtype=fct_relevel(mtype, names(mtypes))) %>% 
ggplot(aes(x=period, y=perc, color=mtype, lty=mtype)) + geom_line() + 
  facet_wrap("party", nrow = 3) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab(NULL) + ylab(NULL) + ggtitle("Visibility of parties per wave", "(as percentage of all political news)") +
  scale_x_continuous(breaks=c(1, 3, 4.5), labels=c("Jan", "Feb", "Mar")) + theme_nrc() +
  theme(legend.title = element_blank(), legend.position = c(1, 0), legend.justification = c("right", "bottom")) +
  scale_linetype_manual(values=c("TV"=1, "Newspapers"=1, "Online (nos.nl/nu.nl)"=1, 
                                 "Social Media (activity)"=2, "Social Media (engagement)"=2))
```

[Download data: `r export_data(hpp2, "Visibility of parties in the media")`]

*TV is INCOMPLETE, IGNORE THE LAST DATAPOINT!*

Overall visibility on traditional media (including online) favours coalition parties VVD, VDA, D66 followed by the main opposition parties.
Over time, Forum become much more visible on these media. PvdA, GL, and SP become  more visible in the newspapers, but less so on TV. 
In contrast, PVV, VVD and FvD (and to a lesser extend D66) are much more visible on television. 

Social media also show a very clear picture: 
although many parties are somewhat active on social media, FvD wins hands down in terms of their activity,
and FvD and PVV are the only parties with substantial engagement on social media (likes, shares, mentions)

# Does the voter take notice?


```{r notice}
d = w %>% filter(variable %in% c("B2", "B3")) %>%  mutate(name=str_remove(name, " *&.*")) %>% 
  mutate(name = case_when(name == "50Plus Partij" ~ "50+",
                          name == "Forum voor Democratie" ~ "FvD",
                          name == "Partij voor de Dieren" ~ "PvdD",
                          T ~ name),
         value = case_when(value == 999 & variable == "B2" ~ 0, 
                           value == 999 & variable == "B3" ~ NA_real_, 
                           T ~ value),
         varlabel=case_when(variable == "B2" ~ "Noticed",
                            T ~ "Positive / Negatibve"),
         name=case_when(name == "ChristenUnie" ~ "CU", 
                        name == "GroenLinks" ~ "GL",
                        name == "50+" ~ "50Plus",
                               T~name))

d2 = d %>% group_by(wave, variable, varlabel, name) %>% summarize(value = mean(value, na.rm=T)) %>%
  mutate(fillcolor = case_when(variable == "B2" ~ 5.5+value,  T ~ value),
         name=fct_rev(factor(name, levels=parties)))

d2 %>% ggplot(aes(x=wave, y=name, fill=fillcolor, label=round(value,1))) + 
  geom_tile(color="white", lwd=1) + facet_grid(cols=vars(varlabel)) + theme_nrc() + 
  scale_fill_gradient2(midpoint=5.5, low=scales::muted("red"), mid="#e0e0e0", high="#01529d", guide=F) + geom_text(color="white") + 
    #scale_fill_gradient2(low=scales::muted("red"), mid="#e0e0e0", high="#01529d", guide=F) + 

  ggtitle("Did respondents notice the parties / leaders?", "(Noticed from 0 (not at all) to 3 (a lot); positive negative from 0 to 10")

```
[Download data: `r export_data(d2, "Media performance according to respondents")`]

Looking at whether respondents noticed the various parties, overall we see that VVD, PVV and FvD are most noticed.
The latter two are interesting since they are less visible in the mainstream media.
So, either they are more noticeable when they are present, or the notice is (also) based on social media presence. 

Most parties are not evaluated particularly positive or negative, with a clear exception for PVV, FvD and Denk,
whose media performance is evaluated very negatively by the respondents. 
This confirms the possibility that PVV and FvD are more noticeable when present, but clearly mostly in a negative way.

Looking at change over time, we see that PvdA and GL become less noticeable, matching their decreasing visibility on TV.
PvdA is evaluated clearly more positively, while VVD, CDA and D66 are evaluated slightly more negatively. 

## Does notice depend on alignment of repondent?

```{r notice-lire}
r2 = resp %>% select(iisID, rile_self, vote) %>% 
  mutate(rile=case_when(is.na(rile_self) ~ NA_character_, rile_self < 5 ~ "links", rile_self > 6 ~ "rechts", T ~ "midden"))

table(w$wave, w$iisID %in% resp$iisID)

w %>% filter(wave == "w2") %>% with(table(iisID))

d3 = d %>% inner_join(r2) %>% filter(!is.na(rile)) %>% group_by(wave, variable, name, rile) %>% summarize(value = mean(value, na.rm=T)) %>% 
  mutate(name=fct_rev(factor(name, levels=parties)))

d3 %>% filter(variable=="B2") %>% 
  ggplot(aes(x=wave, y=name, fill=value, label=round(value,1))) + geom_tile(color="white", lwd=1) + facet_grid(cols=vars(rile)) + theme_nrc() + 
  geom_text(color="white") + scale_fill_gradient(low="#e0e0e0", high="#01529d", guide = F) + 
  ggtitle("Notice of parties by alignment", "(Does respondent self-report as left or right? Values 0 (not noticed) to 3 (noticed a lot))")

```
```{r notice-lire1}
d3 %>% filter(variable=="B3") %>% 
  ggplot(aes(x=wave, y=name, fill=value, label=round(value,1))) + geom_tile(color="white", lwd=1) + facet_grid(cols=vars(rile)) + theme_nrc() + 
  # scale_fill_gradient2(low="red", mid="white", high="green", midpoint = 5.5, guide = F) +
  scale_fill_gradient2(midpoint=5.5, low=scales::muted("red"), mid="#d0d0d0", high="#01529d", guide=F) + geom_text(color="white") + 
  ggtitle("Evaluation of media performence by alignment", "(Does respondent self-report as left or right? Values 0 to 10)")
```

[Download data: `r export_data(d3, "Media performance by alignment")`]

Notice of parties (visibility) does not really depend on the ideological alignment of the respondent.
Evaluation of their media performance does differ: 
leftist voters are overall more critical of rightist parties.
Rightist voters were a bit more critical of GL, SP, D66 and PvdA, but this has mostly converged.


# The Undecided Voter

The number of undecided voters (and non-voters) is becoming smaller as the election draws near:

```{r undecided0}
w0u = votes %>% filter(name == "Undecided", wave == "w0") %>% pull(iisID)
w1u = votes %>% filter(name == "Undecided", wave == "w1") %>% pull(iisID)
w2u = votes %>% filter(name == "Undecided", wave == "w2") %>% pull(iisID)
w2v = votes %>% filter(wave == "w2") %>% pull(iisID)


still_undecided = length(intersect(w0u, w2u)) / length(intersect(w0u, w2v))
newly_undecided = length(setdiff(w2u, w0u)) / length(w2u) 

perc = function(x) glue("{round(x*100)}%")
votes  %>% group_by(wave, name) %>% summarize(n=n())%>% mutate(wave=wave_label(wave), perc=n/sum(n)*100) %>% filter(name %in% c("Undecided", "Not")) %>% select(-n) %>% pivot_wider(names_from=wave, values_from=perc)
```

Very importantly, the group of undecided voters is also changing. 
Of the undecided voters at the start, only `r perc(still_undecided)` is still undecided.
Conversely, of the current undecided voters, `r perc(newly_undecided)` did have a preference at the start. 

## Media use of undecided voters

Last month, we saw that undecided voters were on average less interested in news and politics. 
If we look at the group of voters that were undecided at the start of the campaign, 
how did their media use change?

```{r undecided-media}
d = w %>% mutate(zwevend=fct_rev(ifelse(iisID %in% w0u, "Undecided", "Decided"))) %>% filter(variable == "I1") %>% 
  mutate(name=case_when(name == "Television" ~ "TV",
                        name == "Newspapers or opinion magazines (paper or online)" ~ "Newspapers",
                        name == "Online news sites (such as nu.nl) or blogs" ~ "Online",
                        name == "Social media (such as Facebook, Twitter)" ~ "Social",
                        T ~ "DROP")) %>% 
  filter(name != "DROP") %>% group_by(zwevend, wave, name) %>% summarize(value=mean(value)) 
ggplot(d, aes(x=as.numeric(as.factor(wave)), colour=name, y=value)) + geom_line() + facet_grid(cols=vars(zwevend)) + 
  theme_nrc()  + ggtitle("Media use of *originally* (un)decided voters", "(in frequency of use per week, 0-7)") +
  scale_x_continuous(breaks=c(1,2,3), labels=c("Wave 1\n(Late Jan.)", "Wave 2\n(Early Feb.)", "Wave 3\n(Late Feb.)"))

```

[Download data: `r export_data(d, "Media use of originally (un)decided voters")`]

These results are quite striking: much more than the voters that had already made up their mind,
these undecided voters started consuming more media, both newspapers and TV,
and at the end of February even consume more news. 

```{r undecided-media2}
zwevers = w %>% filter(variable=="vote") %>% mutate(zwevend=fct_rev(ifelse(name=="Undecided", "Undecided", "Decided"))) %>% select(wave, iisID, zwevend)

#d = w %>% filter(variable == "I1") %>% inner_join(zwevers) %>%
d = w %>% filter(iisID %in% c(w2v)) %>% mutate(zwevend=fct_rev(ifelse(iisID %in% w2u, "Undecided", "Decided"))) %>% filter(variable == "I1") %>% 
  mutate(name=case_when(name == "Television" ~ "TV",
                        name == "Newspapers or opinion magazines (paper or online)" ~ "Newspapers",
                        name == "Online news sites (such as nu.nl) or blogs" ~ "Online",
                        name == "Social media (such as Facebook, Twitter)" ~ "Social",
                        T ~ "DROP")) %>% 
  filter(name != "DROP") %>% group_by(zwevend, wave, name) %>% summarize(value=mean(value)) 
ggplot(d, aes(x=as.numeric(as.factor(wave)), colour=name, y=value)) + geom_line() + facet_grid(cols=vars(zwevend)) + 
  theme_nrc()  + ggtitle("Media use of *currently* (un)decided voters", "(in frequency of use per week, 0-7)") +
  scale_x_continuous(breaks=c(1,2,3), labels=c("Wave 1\n(Late Jan.)", "Wave 2\n(Early Feb.)", "Wave 3\n(Late Feb.)"))
```

[Download data: `r export_data(d, "Media use of currently (un)decided voters")`]

If we look at *currently* undecided voters, the picture is strikingly different. 
This group (85% of which did have a preference at the start) actually consume less media across the board,
even as the other respondents are increasing their (political) media use. 

## Who are the undecided voters?

As undecided voters make up their mind, and other voters start doubting their choices, the group of undecided voters changes.
Let's look at some of the demographics:

```{r undecided-demo}
scale = function(x) scales::rescale(as.numeric(x), to=c(0, 100))
r2 = resp %>% mutate(education=scale(education), 
         #age=scale(age), 
         gender=ifelse(gender=="Male", 0, 100),
         leftright=scale(rile_self),
         polknowledge=scale(polknow)) %>% select(iisID, education, age, gender, leftright, polknowledge)

undecideds = w %>% filter(variable == "vote", name=="Undecided") %>% 
  select(wave, iisID) %>% inner_join(r2)
demo_totals = r2 %>% add_column(wave="All respondents")

d = bind_rows(undecideds, demo_totals) %>% 
   group_by(wave) %>% summarize(leftright=mean(leftright, na.rm=T), 
                                polknowledge=mean(polknowledge, na.rm=T), 
                                education=mean(education, na.rm=T), 
                                age=mean(age, na.rm=T), 
                                gender=mean(gender, na.rm=T)) %>%
   pivot_longer(-wave, names_to="measure") %>%
  mutate(wave=fct_relevel(wave, c("w0", "w1", "w2", "All respondents")))

ggplot(d, aes(x=wave, y=measure, fill=value, label=round(value))) +
  geom_tile(lwd=1, color="white") + geom_text(color="white") + theme_nrc() + scale_fill_gradient(guide=F) +
  ggtitle("Demographics of undecided voters per wave", 
          "(polknowledge, education are 0=low, 100=high, gender is percentage female, leftright is 0=left, 100=right)")
```

The table above shows average score on left-right self placement, political knowledge, education, age, and gender.
This shows a clear shift in who those voters are:
The original undecided voters were younger, more often male, higher educated and slighyly more leftist than average.
In constrast, the current undecided voters are centrist and older than the average respondent. 
Interestingly, even though this group has a slightly lower overall education, they score better on political knowledge, probably as a result of being older. 


## Where do undecided voters go to and come from?

```{r undecideds}
parties = c("Undecided", "VVD", "PVV", "CDA", "D66", "GroenLinks", "SP", "PvdA", "ChristenUnie", "PvdD", "50+", "SGP", "Denk", "FvD", "Other", "Not")

#votes %>% filter(name == "Undecided")
w1u = votes %>% filter(name == "Undecided", wave == "w1") %>% pull(iisID)
w2u = votes %>% filter(name == "Undecided", wave == "w2") %>% pull(iisID)
perc_w2 = length(intersect(w1u, w2u)) / length(w1u)

votes = votes %>% na.omit() %>% mutate(waveno=as.numeric(as.factor(wave)), nxt= waveno+1)# %>% mutate(name=ifelse(name %in% c("Not", "Undecided"), "Not/Undecided", name)) 
changes = votes %>% inner_join(votes, by=c("iisID", "nxt"="waveno"), suffix=c("", ".to")) %>% rename(to=name.to) %>% 
  group_by(wave, wave.to, name, to) %>% summarize(n=n())%>% ungroup()

undecided_to = changes %>% filter(wave != "k17", name %in%  c("Not", "Undecided")) %>% group_by(wave, to) %>% summarize(n=sum(n)) #%>% mutate(perc=n/sum(n)*100)%>% select(-n) %>% 
  #pivot_wider(values_fill=0, names_from=wave, values_from=perc) %>% pivot_longer(-to, values_to="perc", names_to="wave")

undecided_from = changes %>% filter(wave != "k17", to %in%  c("Not", "Undecided")) %>% group_by(wave,name) %>% summarize(n=sum(n)) #%>% #mutate(perc=n/sum(n)*100)%>% select(-n) %>% 
  #pivot_wider(values_fill=0, names_from=wave, values_from=perc) %>% pivot_longer(-name, values_to="perc", names_to="wave")

totals = votes %>% group_by(wave) %>% summarize(ntot=n())

d = bind_rows(
  undecided_to %>% rename(party=to) %>% add_column(type="Undecided voters went to"),
  undecided_from %>% rename(party=name) %>% add_column(type="Undecided came from")) %>%
  #tibble(wave="w0", party="Denk", n=0, type="Undecided came from"))  %>% 
  mutate(period=case_when(wave == "w0" ~ "Early February", T ~ "Late February"))

change = d %>% ungroup() %>% select(party, wave, period, type, n) %>% pivot_wider(names_from=type, values_from=n) %>% 
  left_join(totals) %>% 
  mutate(change=`Undecided voters went to` - `Undecided came from`,
         perc=change/ntot*100) %>% 
  select(party, wave, period, n=change, perc) %>% add_column(type="Net Change")


d2 = bind_rows(d, change) 
d2 = d2 %>% left_join(totals) %>% filter(party!="Undecided") %>% mutate(
  perc = n/ntot*150,
  colval=ifelse(type=="Net Change", perc*3, perc), 
  type=fct_rev(type),
  party2=fct_rev(factor(party, levels=parties)))

ggplot(d2, aes(x=period, y=party, fill=colval, label=round(perc, 1))) + geom_tile(color="white", lwd=2) + geom_text(color="white") +
  facet_grid(cols=vars(type)) +
  scale_fill_gradient2(low=scales::muted("red"), mid="#e0e0e0", high="#01529d", guide=F) + 
  xlab(NULL) + ylab(NULL) + ggtitle("Where do undecided voters go to and come from?", "(as virtual seats)") +
  ggthemes::theme_hc() + theme(panel.grid.major.y = element_blank(), strip.background = element_rect(fill="white"))
```

[Download data: `r export_data(d, "Undecided Voters")`]

On the left hand side, the figure above shows where undecided voters go to when they make up their mind.
Similarly, the middle columns show where new undecided voters go to. 
Finally, the rightmost two columns show the difference between the two: 
was a party able to attract more undecided voters they lost? 

The first thing that is very striking is that many party both attract a substantial number of undecided voters, while at the same time losing voters that become undecided. 
In fact, only `r perc_w2`% of undecided voters in wave 2 were also undecided in wave 1!
A lot of respondents switch in and out of the undecided group. 

Looking at which parties gain or lose from undecided voters, overall most parties make net gains as the number of undecided voters goes down.
VVD, Groenlinks and PvdA and CU are mostly able to profit in early February, with the latter three even making a net loss later that month.
Apparently, many voters started doubting their choice for these parties.
In contrast, CDA and D66 but also PvdD and FvD actually profited most in the second half of February. 

